Thesis Title

### 2022-10-17 chapters/discussion.tex
copy-editing on discussion

Metagenomics, and in particular \ac{SMg}, has emerged as a promising approach for diagnosis from clinical samples and surveillance of organisms of interest from the environment \citep{loman_culture-independent_2013, rossen__2018, schuele_future_2021, chiu_clinical_2019}. A single metagenomics analysis has the potential to detect common, rare and novel pathogens, and provides a broad overall picture of the microbial content present in a sample. Despite this, it is often unclear whether a given detected microorganism is a contaminant, coloniser or \textit{bona fide} pathogen.

@@

Second generation sequencing so far represent the most popular technology applied in metagenomics \citep{rossen_practical_2018, loman_twenty_2015, loman_high-throughput_2012, schuele_future_2021}. Second-generation methods require library preparation including an enrichment or amplification step \citep{hagemann_overview_2015}, a time-consuming, bias-inducing procedure that is propagated to the resulting data. Another limitation is the size of the outputted sequences that, despite the massive throughput of some of the machines available, requiring a very small \ac{DNA} input load to produce up to billions of sequences \citep{loman_twenty_2015}, ranges from 45 to 300 bases in length \citep{loman_performance_2012, ari_next-generation_2016}. This is simply not enough to transverse the most repetitive genomic regions, and severely limits the sensitivity of the methodology as the source organism of the short sequences is hard to determine both by mapping and \textit{de novo} assembly methods. Furthermore, turnaround times range from several hours to full days, not ideal for a timely diagnosis and reporting, and require batching of samples to be run cost-effectively which, depending on the instrument, can range from a few samples to a few dozens, adding to the turnaround time and lowering the sensitivity of pathogen detection \citep{greninger_challenge_2018, schuele_future_2021}. In Chapter \ref{ch:paper1}, the analysis through \ac{SMg} of 10 samples took between 48-54 hours to complete, which is shorter than culture-based methods if one includes typing. Due to the sequencing depth required to capture the lowest abundant microorganisms in a sample, only the instruments with the highest throughput are recommended for metagenomic sequencing \footnote{ \url{https://www.illumina.com/systems/sequencing-platforms.html}}. This, however, is not cost-effective in routine diagnostics where samples need to be immediately processed \citep{rossen__2018}. Also, as discussed in Chapter \ref{ch:paper1}, when applying sample batching, second generation sequencing-related phenomena such as index hopping (also named index switching) or crosstalk (also called sample bleeding) can introduce false-positive results.

Third generation sequencing technologies have emerged as a viable alternative to second generation methods as longer sequences offer more contextual information and do not have the limitation of bias-inducing pre-sequencing \ac{PCR} library preparation, as each molecule is sequenced directly \citep{loman_twenty_2015, ari_next-generation_2016}. This results in a lower bias, but a much lower throughput of data and a much higher baseline sample input requirement \citep{gu_clinical_2019}. Long-reads also have the advantage of resolving structural variations and variants in repetitive regions, which are poorly resolved by short-reads and are often excluded in bioinformatics analysis. In particular, \ac{ONT} sequencing has emerged as an attractive platform for clinical laboratories to adopt due to its low cost and rapid turnaround time, being able to provide results in almost real-time \citep{petersen_third-generation_2019}. Despite this, this method still faces problems in base-calling accuracy when compared with other platforms \citep{gu_clinical_2019}. In Chapter \ref{ch:paper2} it was shown that even when hybrid assembly with both Illumina and \ac{ONT} data is employed, leveraging second and third generation methodologies, complete genomic sequences, particularly chimeric ones such as plasmids, are still not fully recovered. Although Nanopore sequencing is able to produce reads of up to 2 megabases in length, the biggest drawbacks to date have been a lower throughput of sequence data and a high error rate (approximately 10\%) \citep{petersen_third-generation_2019}. Regardless, rapid advancements are being achieved with new, more accurate flowcells and better base calling software \citep{sereika_oxford_2021}.

@@

The unbiased nature of \ac{SMg} allows the sequencing of the nucleic acid of all microbes (including commensal) and the host. Nearly all \ac{DNA} and \ac{RNA} content in most clinical samples is host (human) derived, further lowering the sensitivity of potential pathogen detection \citep{gu_clinical_2019}. Therefore, the presence of an overwhelming amount of host \ac{DNA} or \ac{RNA} is one of the most important problems to be addressed in \ac{SMg}.
As observed in Chapter \ref{ch:paper1}, the number of human reads differed between the 10 samples selected form \ac{SMg}, even when using the same extraction kit, and all kits including a human DNA depletion step. This highlights that the ratio between host and microbial DNA is a major determinant of the proportion, and therefore the detection of microbial reads.

There are depletion steps that aim to decrease the relative proportion of human host background sequences through capture probes, lysis and deoxyribonuclease and/or ribonuclease treatment \citep{gu_clinical_2019}. Theoretically, the microbial proportion of the sample is protected within viral capsids and microbial cell walls, but alterations to the microbial composition can still occur. Alternatively, the host sequences can be removed after sequencing, as performed in Chapter \ref{ch:paper3}. In this Chapter, the reads of interest were captured through a mapping approach to a large collection of reference genomes but the opposite methodology can be applied, where the contaminant host sequences are removed (in their majority) by mapping to a human reference genome. This method is not as cost-efficient as host DNA depletion in the bench, as a greater proportion of background sequences are sequenced, but the community composition remains, theoretically, unaltered.

@@

In addition to the variability of wet lab protocols, the bioinformatics for data handling and interpretation can be resource intensive. Bioinformatic analysis requires highly trained staff, valid analysis tools, including the reference database, good computational infrastructure, and the creation of standardised procedures \citep{petersen_third-generation_2019}. In routine settings, automation and standardisation of the analysis are significant for the reliability of the diagnostic and surveillance test results. Computational pipelines for the analysis of \ac{SMg} have unique challenges and requirements, from host depletion and taxonomic classification, to metagenomic \textit{de novo} assembly, binning of \ac{MAG}s and strain characterisation. Additionally, as observed in Chapters \ref{ch:paper1} and \ref{ch:paper2}, each sequencing platform, from second (Illumina) and third generation (\ac{ONT}), requires their own data processing steps and quality control metrics as the tools that have been developed in the research community for short-read data are not feasible for long-read data. Data interpretation adds an additional level of complexity.

For \textit{in silico} host depletion, only the sequences that align to a human reference genome will be removed, not necessarily removing the totality of human sequences and, in the lens of patient identity protection, leaving the most identifiable sequences behind. In the same principle, when performing taxonomic characterisation, only sequences present in the database used for sequence identification will be identified, with rare pathogens or emerging strains of pathogens not being reported. In Chapter \ref{ch:paper1}, different taxonomic classification tools produced different results for the same sample. Additionally, reference databases are often biased towards certain organisms, as well as certain pathogens that are important to distinguish are similar genetically, with current methods not having enough resolution to do so \citep{gu_clinical_2019}. Lastly, contamination with flora and/or reagents should be accounted for as it can limit specificity, hence the importance of using and analysing controls alongside samples.

The quality control of the \ac{SMg} reads is in all similar to any genomics workflow, an essential prerequisite that involves quality trimming and sequencing adaptor removal. In opposition, the assembly step, where reads are stitched into longer fragments, referred to as contigs, is usually tailored to \ac{SMg} analysis. These contigs are longer sequences that offer better contextual information than reads alone and provide a more complete picture of the microbial community than just the species composition and are usually followed by reconstructing the individual genes and species. In Chapter \ref{ch:paper4} both metagenomic and traditional genomic assemblers are employed in a two-pronged approach to guarantee the highest chance of success when recovering full \ac{DENV} genomes from the metagenomic samples. As discussed in Chapter \ref{ch:paper5}, several dedicated metagenomic assembly tools for short-read data are available, generally assumed to perform better when dealing with the complex \ac{SMg} samples. Despite this assumption, assemblers branded as metagenomic specific did not consistently outperform other genomic assemblers in the metagenomic samples analysed. Additionally, the performance of each assembler varied depending on the species of interest and its abundance in the sample, with less abundant species presenting a significant challenge for all assemblers.

@@

In Chapter \ref{ch:paper4}, where ultimately a two-pronged approach was followed including both a genomic and a metagenomic \textit{de novo} assembler. This arose from the need to recover complete CDS sequences, where none of the alternatives on their own was able to perform consistently. Despite several benchmarking studies being available \citep{miller_assembly_2010, earl_assemblathon_2011, bradnam_assemblathon_2013, liao_current_2019, chen_benchmarking_2020, sczyrba_critical_2017, meyer_critical_2022}, the assemblers benchmarked and the datasets used are not consistent, usually limited to a couple of different tools that do not reflect the current landscape of possible \textit{de novo} assembly software. These benchmarks are also usually coupled with the release of a new tool, therefore often skewed towards certain results and do not offer an unbiased assessment.

In Chapter \ref{ch:paper5}, 12 \textit{de novo} assemblers were benchmarked with the same metagenomic samples, selected based on the date of the last update to the software and with a cut-off in 2015. No assembler stood out as an undisputed all-purpose choice for short-read metagenomic prokaryote \textit{de novo} genome assembly, highlighting that efforts are still needed to further improve metagenomic assembler performance.

@@

Chapter \ref{ch:paper5} shows that suitable mock communities, reproducing the users’ samples of interest, can be used as a gold standard to evaluate tool performance. Several well-characterised mock communities are currently available to be used by the community when assessing and benchmarking software suitable for \ac{SMg} analysis. Chapter \ref{ch:paper5} features the ZymoBIOMICS Microbial Community Standard, composed of eight bacterial species and two fungi, and is available commercially, with reference sequences made publicly available, with both even and logarithmically distribution of species\footnote{\url{https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards}}. Still, ZymoBIOMICS Microbial Community Standards might not be representative of the metagenomic complexity of the samples of interest of most researchers, its relative simplicity means that the results shown probably represent a best-case scenario. Furthermore, in the logarithmically distributed sample, six replicons have a coverage lower than one time, rendering it unrecoverable through assembly. To address this, two other community standards were analysed: the BMock12 \citep{sevim_shotgun_2019} and the Gut-Mix-RR \citep{amos_developing_2020}.

@@

The Gut-Mix-RR community dataset, and its uneven distribution of species counterpart Gut-HiLo-RR, spans five phyla, thirteen families, sixteen genera and nineteen species, comprising twenty strains that aim to emulate the healthy human gut microbiota. The strains were obtained from Leibniz Institute DSMZ-German Collection of Microorganisms and Cell Cultures GmbH. The genomes in the mock community are fairly diverse, with the challenging genomes being the two \textit{Bifidobacterium longum} subspecies (ANIb=0.95). It’s worth denoting that no complete genome is available for eight of the genomes, including one of the \textit{Bifidobacterium longum} subspecies. This makes the assessment of results challenging, if not impossible, as no source of absolute truth is available.

For the replication of more complex communities, the MICROBIOME Community of Special Interest\footnote{\url{https://www.microbiome-cosi.org/}} makes available, through their \ac{CAMI} initiative\footnote{\url{https://data.cami-challenge.org/}}, several datasets varying in complexity, replicating several communities such as human microbiome and the rhizosphere \citep{sczyrba_critical_2017, meyer_critical_2022}. Unfortunately, sources of truth are not made available, including the complete genomes of the community profile, and assessment of results is difficult, if not impossible. Despite their popularity, no information is provided in the manuscripts or in the supplemental materials about the source of the genomes in each dataset. It is only mentioned that the community composition was designed according to specified criteria. Additionally, only assemblies are made available, and not the source references for the generation of the mock datasets. These assemblies are, unfortunately, often the result of a pooled assembly and are extremely fragmented, with N50 ranging from 18406 basepairs in 56168 contigs to 58988 basepairs in 508968 contigs. This makes this dataset not suitable for benchmark studies, which require defined mock communities as a source of truth.

@@

No ideal solution is currently available in the research community or the market, highlighting the need for better, more diverse and well-described community standards representing realistic scenarios.

>>>>

Metagenomics, and in particular \ac{SMg}, has emerged as a promising approach for diagnosis from clinical samples and surveillance of organisms of interest from the environment \citep{loman_culture-independent_2013, rossen__2018, schuele_future_2021, chiu_clinical_2019}. A single metagenomics analysis has the potential to detect common, rare and novel pathogens, and provide a broad overall picture of the microbial content present in a sample. Despite this, it is often unclear whether a given detected microorganism is a contaminant, coloniser or \textit{bona fide} pathogen.

@@

Second generation sequencing so far represent the most popular technology applied in metagenomics \citep{rossen_practical_2018, loman_twenty_2015, loman_high-throughput_2012, schuele_future_2021}. Second-generation methods require library preparation including an enrichment or amplification step \citep{hagemann_overview_2015}, a time-consuming, bias-inducing procedure that is propagated to the resulting data. Another limitation is the size of the outputted sequences that, despite the massive throughput of some of the machines available, requiring a very small \ac{DNA} input load to produce up to billions of sequences \citep{loman_twenty_2015}, ranges from 45 to 300 bases in length \citep{loman_performance_2012, ari_next-generation_2016}. This is simply not enough to transverse the most repetitive genomic regions, and severely limits the sensitivity of the methodology as the source organism of the short sequences is hard to determine both by mapping and \textit{de novo} assembly methods. Furthermore, turnaround times range from several hours to full days, not ideal for a timely diagnosis and reporting, and require batching of samples to be run cost-effectively which, depending on the instrument, can range from a few samples to a few dozen, adding to the turnaround time and lowering the sensitivity of pathogen detection \citep{greninger_challenge_2018, schuele_future_2021}. In Chapter \ref{ch:paper1}, the analysis through \ac{SMg} of 10 samples took between 48-54 hours to complete, which is shorter than culture-based methods if one includes typing. Due to the sequencing depth required to capture the lowest abundant microorganisms in a sample, only the instruments with the highest throughput are recommended for metagenomic sequencing \footnote{ \url{https://www.illumina.com/systems/sequencing-platforms.html}}. This, however, is not cost-effective in routine diagnostics where samples need to be immediately processed \citep{rossen__2018}. Also, as discussed in Chapter \ref{ch:paper1}, when applying sample batching, second generation sequencing-related phenomena such as index hopping (also named index switching) or crosstalk (also called sample bleeding) can introduce false-positive results.

Third generation sequencing technologies have emerged as a viable alternative to second generation methods as longer sequences offer more contextual information and do not have the limitation of bias-inducing pre-sequencing \ac{PCR} library preparation, as each molecule is sequenced directly \citep{loman_twenty_2015, ari_next-generation_2016}. This results in a lower bias, but a much lower throughput of data and a much higher baseline sample input requirement \citep{gu_clinical_2019}. Long-reads also have the advantage of resolving structural variations and variants in repetitive regions, which are poorly resolved by short-reads and are often excluded in bioinformatics analysis. In particular, \ac{ONT} sequencing has emerged as an attractive platform for clinical laboratories to adopt due to its low cost and rapid turnaround time, being able to provide results in almost real-time \citep{petersen_third-generation_2019}. However, compared to other platforms \citep{gu_clinical_2019}, this method still has problems with base-calling accuracy. Although it can produce reads of up to 2 megabases in length, its main drawbacks have been a low throughput of sequence data and a high error rate (approximately 10\%). In Chapter \ref{ch:paper2} it was shown that even when hybrid assembly with both Illumina and \ac{ONT} data is employed, leveraging second and third generation methodologies, complete genomic sequences, particularly chimeric ones such as plasmids, are still not fully recovered. Regardless, rapid advancements are being achieved with new, more accurate flowcells and better base calling software \citep{sereika_oxford_2021}.

@@

The unbiased nature of \ac{SMg} allows the sequencing of the nucleic acid of all microbes (including commensal) and the host. Nearly all \ac{DNA} and \ac{RNA} content in most clinical samples is host (human) derived, further lowering the sensitivity of potential pathogen detection \citep{gu_clinical_2019}. As a result, the presence of an overwhelming amount of host \ac{DNA} or \ac{RNA} is one of the most important problems to be addressed in \ac{SMg}.
As observed in Chapter \ref{ch:paper1}, the number of human reads differed between the 10 samples selected for \ac{SMg}, even when using the same extraction kit and including a human DNA depletion step. This highlights that the ratio between host and microbial DNA is a major determinant of the proportion, and therefore the detection of microbial reads.

There are depletion steps that aim to decrease the relative proportion of human host background sequences through capture probes, lysis and deoxyribonuclease and/or ribonuclease treatment \citep{gu_clinical_2019}. Theoretically, the microbial proportion of the sample is protected within viral capsids and microbial cell walls, but alterations to the microbial composition can still occur. Alternatively, the host sequences can be removed after sequencing, as performed in Chapter \ref{ch:paper3}. In this Chapter, the reads of interest were captured through a mapping approach to a large collection of reference genomes. Still, the opposite methodology can be applied, where the contaminant host sequences are removed (in their majority) by mapping to a human reference genome. This method is not as cost-efficient as host DNA depletion in the bench, as a greater proportion of background sequences are sequenced, but the community composition remains, theoretically, unaltered.

@@

In addition to the variability of wet lab protocols, the bioinformatics for data handling and interpretation can be resource intensive. Bioinformatic analysis requires highly trained staff, valid analysis tools, including the reference database, good computational infrastructure, and the creation of standardised procedures \citep{petersen_third-generation_2019}. In routine settings, automation and standardisation of the analysis are significant for the reliability of the diagnostic and surveillance test results. Computational pipelines for the analysis of \ac{SMg} data have unique challenges and requirements, from host depletion and taxonomic classification, to metagenomic \textit{de novo} assembly, binning of \ac{MAG}s and strain characterisation. Additionally, as observed in Chapters \ref{ch:paper1} and \ref{ch:paper2}, each sequencing platform, from second (Illumina) and third generation (\ac{ONT}), requires their own data processing steps and quality control metrics as the tools that have been developed in the research community for short-read data are not feasible for long-read data. Data interpretation adds an additional level of complexity.

For \textit{in silico} host depletion, only the sequences that align to a human reference genome will be removed, not necessarily removing the totality of human sequences and, in the lens of patient identity protection, leaving the most identifiable sequences behind. In the same principle, when performing taxonomic characterisation, only sequences present in the database used for sequence identification will be identified, with rare pathogens or emerging strains of pathogens not being reported. This is a limitation of DEN-IM, presented in Chapter \ref{ch:paper3}, where the latter approach is implemented. In Chapter \ref{ch:paper1}, different taxonomic classification tools produced different results for the same sample. Additionally, reference databases are often biased towards certain organisms, as well as certain pathogens that are important to distinguish are similar genetically, with current methods not having enough resolution to do so \citep{gu_clinical_2019}. Lastly, contamination with flora and/or reagents should be accounted for as it can limit specificity, hence the importance of using and analysing controls alongside samples.

The quality control of the \ac{SMg} reads is in all similar to any genomics workflow, an essential prerequisite that involves quality trimming and sequencing adaptor removal. In opposition, the assembly step, where reads are stitched into longer fragments, referred to as contigs, is usually tailored to \ac{SMg} analysis. These contigs are longer sequences that offer better contextual information than reads alone and provide a more complete picture of the microbial community than just the species composition and are usually followed by reconstructing the individual genes and species. In Chapter \ref{ch:paper3} both metagenomic and traditional genomic assemblers are employed in a two-pronged approach to guarantee the highest chance of success when recovering full \ac{DENV} genomes from the metagenomic samples. As discussed in Chapter \ref{ch:paper4}, several dedicated metagenomic assembly tools for short-read data are available, generally assumed to perform better when dealing with the complex \ac{SMg} samples. Despite this assumption, assemblers branded as metagenomic specific did not consistently outperform other genomic assemblers in the metagenomic samples analysed. Additionally, the performance of each assembler varied depending on the species of interest and its abundance in the sample, with less abundant species presenting a significant challenge for all assemblers.

@@

In Chapter \ref{ch:paper3}, where ultimately a two-pronged approach was followed including both a genomic and a metagenomic \textit{de novo} assembler. This arose from the need to recover complete CDS sequences, where none of the alternatives on their own was able to perform consistently. Despite several benchmarking studies being available \citep{miller_assembly_2010, earl_assemblathon_2011, bradnam_assemblathon_2013, liao_current_2019, chen_benchmarking_2020, sczyrba_critical_2017, meyer_critical_2022}, the assemblers benchmarked and the datasets used are not consistent, usually limited to a couple of different tools that do not reflect the current landscape of possible \textit{de novo} assembly software. These benchmarks are also usually coupled with the release of a new tool, therefore often skewed towards certain results and do not offer an unbiased assessment.

In Chapter \ref{ch:paper4}, 11 \textit{de novo} assemblers were benchmarked with the same metagenomic samples, selected based on the date of the last update to the software and with a cut-off in 2015. No assembler stood out as an undisputed all-purpose choice for short-read metagenomic prokaryote \textit{de novo} genome assembly, highlighting that efforts are still needed to further improve metagenomic assembler performance.

@@

Chapter \ref{ch:paper4} shows that suitable mock communities, reproducing the users’ samples of interest, can be used as a gold standard to evaluate tool performance. Several well-characterised mock communities are currently available to be used by the community when assessing and benchmarking software suitable for \ac{SMg} analysis. Chapter \ref{ch:paper5} features the ZymoBIOMICS Microbial Community Standard, composed of eight bacterial species and two fungi, and is available commercially, with reference sequences made publicly available, with both even and logarithmically distribution of species\footnote{\url{https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards}}. Still, ZymoBIOMICS Microbial Community Standards might not be representative of the metagenomic complexity of the samples of interest of most researchers, its relative simplicity means that the results shown probably represent a best-case scenario. Furthermore, in the logarithmically distributed sample, six replicons have a coverage lower than one time, rendering it unrecoverable through assembly and possibly mapping. To address this, two other community standards were analysed: the BMock12 \citep{sevim_shotgun_2019} and the NIBSC Gut DNA Reference \citep{amos_developing_2020}.

@@

The NIBSC Gut DNA Reference community dataset, with the evenly distributed Gut-Mix-RR sample and its uneven distribution of species counterpart Gut-HiLo-RR, spans five phyla, thirteen families, sixteen genera and nineteen species, comprising twenty strains that aim to emulate the healthy human gut microbiota. The strains were obtained from Leibniz Institute DSMZ-German Collection of Microorganisms and Cell Cultures GmbH. The genomes in the mock community are fairly diverse, with the challenging genomes being the two \textit{Bifidobacterium longum} subspecies (ANIb=0.95). It’s worth denoting that no complete genome is available for eight of the genomes, including one of the \textit{Bifidobacterium longum} subspecies. This makes the assessment of results challenging, if not impossible, as no source of absolute truth is available.

For the replication of more complex communities, the MICROBIOME Community of Special Interest\footnote{\url{https://www.microbiome-cosi.org/}} makes available, through their \ac{CAMI} initiative\footnote{\url{https://data.cami-challenge.org/}}, several datasets varying in complexity, replicating several communities such as human microbiome and the rhizosphere \citep{sczyrba_critical_2017, meyer_critical_2022}. Unfortunately, sources of truth are not made available, including the complete genomes of the community profile, and assessment of results is difficult, if not impossible. Despite their popularity, no information is provided in the manuscripts or in the supplemental materials about the source of the genomes in each dataset. It is only mentioned that the community composition was designed according to undisclosed specified criteria. Additionally, only assemblies are made available, and not the source references for the generation of the mock datasets. These assemblies are, unfortunately, often the result of a pooled assembly and are extremely fragmented. This makes this dataset not suitable for benchmark studies, which require defined mock communities as a source of truth.

@@

No ideal solution is currently available in the research community or the market, highlighting the need for better, more diverse and well-described community standards representing realistic scenarios. Chapter \ref{ch:paper4} the results of the ZymoBIOMICS Microbial Community Standard, BMock12 and NIBSC Gut DNA community standards suggest that the actual genome of interest and community composition play an important part in the results of individual assemblers. Particularly, The performance of each assembler varied depending on the species of interest and its abundance in the sample. For this reason, users should analyse the results of sequencing mock communities (ideally) or artificially generated reads simulating their samples of interest to guide their downstream metagenomic analysis and tool selection.
